Disentangling linkages between satellite-derived indicators of forest structure and productivity for ecosystem monitoring.

### 2024-01-10 scripts/manuscript/manuscript.qmd
began revisions round 2, cleaned 4_vegan to be easier to work with (less defunct extra code)

---
title: Disentangling linkages between satellite-derived indicators of forest structure and productivity for ecosystem monitoring.
author:
  - name: Evan R. Muise
    email: evan.muise@student.ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
    attributes:
        corresponding: true
  - name: Margaret E. Andrew
    email: M.Andrew@murdoch.edu.au
    affiliations: 
        - id: murd
          name: Murdoch University
          department: Environmental and Conservation Sciences and Harry Butler Institute
          address: 90 South St
          city: Murdoch, WA, Australia
          postal-code: 6150
  - name: Nicholas C. Coops
    email: nicholas.coops@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
  - name: Txomin Hermosilla
    email: txomin.hermosillagomez@NRCan-RNCan.gc.ca
    affiliations: 
        - id: cfs
          name: Natural Resources Canada
          department: Canada Forest Service (Pacific Forestry Centre)
          address: 506 Burnside Rd W
          city: Victoria, BC, Canada
          postal-code: V8Z 1M5
  - name: A. Cole Burton
    email: cole.burton@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Stephen S. Ban
    email: Stephen.Ban@gov.bc.ca
    affiliations: 
        - id: bcp
          name: Ministry of Environment and Climate Change Strategy
          department: BC Parks
          address: 525 Superior Street
          city: Victoria, BC, Canada
          postal-code: V8V 1T7
keywords: 
  - remote sensing
  - landsat
  - forest structure
  - forest productivity
  - dynamic habitat indices (DHIs)
  - essential biodiversity variables
  - redundancy analysis (RDA)
date: last-modified
bibliography: [bibliography.bib, packages.bib]
csl: scientific-reports.csl
format:
  elsevier-pdf:
    keep-tex: true
    keep-md: true
    journal:
      name: Nature Scientific Reports
      formatting: review
      model: 3p
      cite-style: authoryear
      layout: twocolumn
  docx: 
    reference-doc: my-styles.docx
    keep-md: true
  pdf: 
    pdf-engine: xelatex
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: false
fig-width: 300
number-sections: true
nocite: |
  @matasci2018, @matasci2018b
---

@@

The goal of this study is to assess the complementarity of two remote
sensing datasets often proposed as indicators of ecosystem structure and
function EBVs. We interpret weak or absent relationships between metrics
to constitute complementary information. To do so, we synthesize
wall-to-wall forest structure data previously generated by modelling
lidar-derived forest structure metrics with medium spatial resolution
satellite imagery using a k-NN imputation method [@matasci2018;
@matasci2018b], alongside a well-established remote sensing-derived
index of ecosystem function (DHIs) [@coops2008; @berry2007;
@radeloff2019; @razenkovaMediumresolutionDynamicHabitatInPress]. First,
we analyze remote sensing indicator complementarity by examining the
multivariate relationship between ecosystem structure and function.
Second, we propose a method for selecting which remote sensing
attributes to develop and apply within a single EBV class by determining
the independent and shared contributions of lidar-derived forest
structural attributes and modelled forest structure to explaining
ecosystem function. Third, we examine the global applicability of these
results by evaluating the consistency of the observed forest
structure-function relationships across a broad range of environment
gradients and forest types.

@@

```{r}
#| label: fig-study-map
#| fig-cap: Location of Biogeoclimatic Ecosystem Classification (BEC) zones in British Columbia.

@@

bec_tab %>%
  gt() %>%
  cols_label(
    group = "BEC Group",
    zone = "BEC Zone",
    zone_nm = "Full Name",
    ppt = "Average Monthly Precipitation (mm/yr)",
    tmax = "Average Monthly Max Temperature (°C)",
    tmin = "Average Monthly Min Temperature (°C)"
  ) %>%
  tab_header(title = "Biogeoclimatic Ecosystem Classification (BEC) Zones, their aggregated biomes, and their monthly average climate values for precipitation, maximum temperature, and minimum temperature. Climate data from Wang et al. 2016 averaged for 1991-2020.")
```

@@

The forest structural attributes (canopy height, canopy cover,
structural complexity \[coefficient of variation of height returns\],
basal area, aboveground biomass, and gross stem volume) were derived
from the 30-m spatial resolution dataset generated by Matasci et al.
[@matasci2018; @matasci2018b]. In brief, the method extends structural
attributes derived from a set of lidar acquisitions and field plots
across Canada. These data were generated for the treed pixels of Canada
using the BAP Landsat surface reflectance composite and auxiliary data
(i.e., topography, geography) using a k-Nearest Neighbour imputation
[@matasci2018; @matasci2018b]. Accuracy metrics for imputed forest
structural attributes range from an RMSE of 24.5% (structural
complexity) to 82.3% (gross stem volume) and *R^2^* ranging from 0.125
(structural complexity) to 0.712 (gross stem volume) [@matasci2018;
@matasci2018b].

@@

To obtain representative sample units of the EBV metrics across the
study area, we implemented stratified random sampling across the union
of BEC zones (Table 1) and forest types (i.e., coniferous, broadleaf,
mixed wood, and wetland treed; see @sec-study-area) of BC. Sample units
were restricted to forested pixels that were surrounded by the same
forest type to reduce the chance of sampling mixed pixels. We also
required that sample units were selected from a homogeneous area, to
reduce uncertainty associated with the input datasets [@shang2020];
homogeneous areas were defined as 3x3-pixel windows with a coefficient
of variation for canopy cover and canopy height below 0.5. A minimum
sampling distance of 1-km was implemented to reduce the effects of
spatial autocorrelation. Sample units that had been disturbed in the
last 35 years were discarded by using the disturbance mask generated for
the forested ecosystems of Canada [@hermosilla2016]. A maximum of 500
sample units were selected from each stratum; in the many strata where
this was not achieved, all sampling units meeting the above restrictions
were selected. Sampling was conducted in R version 4.2.2 [@R-base] using
the **sgsR** package [@R-sgsR]. Neighbourhood analyses for the land
cover classes and coefficient of variations of canopy height and cover
were calculated in Python version 3.9.

@@

```{r}
#| label: fig-fcc
#| fig-cap: False colour maps by Biogeoclimatic Ecosystem Classification (BEC) zone of axis loadings for the first redundancy analysis (RDA) axis (A and B) and second RDA axis (C and D). Colour values are normalized to the maximum loading of each variable. A and C show axis loadings for canopy height (CH; red), canopy cover (CC; green) and structural complexity (SC; blue). B and D show axis loadings for basal area (BA; red), total aboveground biomass (AB; green) and gross stem volume (GSV; blue). Black indicates no significant loadings, while greyscale colours indicate similar loadings across the three variables. Annotated colour wheels for each set of variables also shown.

@@

BC shows varying forest structural drivers of the DHIs depending on the
geographic location (@fig-fcc). The interior of BC is drier and warmer,
and also has productivity being strongly driven by canopy cover, while
the Coastal BEC zone group has very low canopy cover loadings and high
structural complexity loadings (@fig-radar). The climate of these
regions could be influencing the DHIs [@perez2016], leading to varying
loadings strengths depending on the distance to the coast. For example,
water availability has been shown to be a driver of canopy height
[@klein2015]. In these wetter regions such as the Coastal group,
structural variability is more likely to be present in the vertical,
rather than horizontal direction, especially given the relationship
between canopy height and structural complexity [@atkins2022]. Further,
the weakest relationships between structure and function were found in
the Coastal group of BEC zones (@fig-fracts-bar), which may be due to
the consistently high amounts of canopy cover, and lower amounts of snow
found in these zones leading to potentially saturated NDVI values
throughout the year. These consistent forest structural metrics --- such
as the high canopy cover values found in the Coastal BEC zones --- may
also lead to lower amounts of variation explained by these metrics,
leading to them being less informative in the loadings, and other
metrics, such as structural complexity having higher amounts of variance
explained (@fig-radar). These complex linkages between climate, forest
structure, and forest productivity become further complicated by the
extreme environmental and topographic gradients within BC's ecosystems
[@pojar1987; @valentine1978soil], which may warrant study into the scale
of these interactions, with a finer (or non-climate based)
ecoregionalization potentially revealing different relationships between
ecosystem structure and function.

>>>>

---
title: Disentangling linkages between satellite-derived indicators of forest structure and productivity for ecosystem monitoring.
author:
  - name: Evan R. Muise
    email: evan.muise@student.ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
    attributes:
        corresponding: true
  - name: Margaret E. Andrew
    email: M.Andrew@murdoch.edu.au
    affiliations: 
        - id: murd
          name: Murdoch University
          department: Environmental and Conservation Sciences and Harry Butler Institute
          address: 90 South St
          city: Murdoch, WA, Australia
          postal-code: 6150
  - name: Nicholas C. Coops
    email: nicholas.coops@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
  - name: Txomin Hermosilla
    email: txomin.hermosillagomez@NRCan-RNCan.gc.ca
    affiliations: 
        - id: cfs
          name: Natural Resources Canada
          department: Canada Forest Service (Pacific Forestry Centre)
          address: 506 Burnside Rd W
          city: Victoria, BC, Canada
          postal-code: V8Z 1M5
  - name: A. Cole Burton
    email: cole.burton@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Stephen S. Ban
    email: Stephen.Ban@gov.bc.ca
    affiliations: 
        - id: bcp
          name: Ministry of Environment and Climate Change Strategy
          department: BC Parks
          address: 525 Superior Street
          city: Victoria, BC, Canada
          postal-code: V8V 1T7
keywords: 
  - remote sensing
  - landsat
  - forest structure
  - forest productivity
  - dynamic habitat indices (DHIs)
  - essential biodiversity variables
  - redundancy analysis (RDA)
date: last-modified
bibliography: [bibliography.bib, packages.bib]
csl: scientific-reports.csl
format:
  elsevier-pdf:
    keep-tex: true
    keep-md: true
    journal:
      name: Nature Scientific Reports
      formatting: review
      model: 3p
      cite-style: authoryear
      layout: twocolumn
  docx: 
    reference-doc: my-styles.docx
    keep-md: true
  pdf: 
    pdf-engine: xelatex
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: false
fig-width: 300
number-sections: true
nocite: |
  @R-ggplot2, @sf2023
---

@@

The goal of this study is to assess the complementarity of two remote
sensing datasets often proposed as indicators of ecosystem structure and
function EBVs. We interpret weak or absent relationships between metrics
to constitute complementary information. To do so, we synthesize
wall-to-wall forest structure data previously generated by modelling
lidar-derived forest structure metrics with medium spatial resolution
satellite imagery using a k-NN imputation method [@matasci2018;
@matasci2018b], alongside a well-established remote sensing-derived
index of ecosystem function (DHIs) [@coops2008; @berry2007;
@radeloff2019; @razenkovaMediumresolutionDynamicHabitatInPress]. First,
we analyze remote sensing indicator complementarity by examining the
multivariate relationship between ecosystem structure and function.
Second, we propose a method for selecting which remote sensing
attributes to develop and apply within a single EBV class by determining
the independent and shared contributions of lidar-derived forest
structural attributes and modelled forest structure to explaining
ecosystem function. Third, we examine the potential geographic
extensibility of these results by evaluating the consistency of the
observed forest structure-function relationships across a broad range of
environment gradients and forest types.

@@

```{r}
#| label: fig-study-map
#| fig-cap: Location of Biogeoclimatic Ecosystem Classification (BEC) zones in British Columbia. Map was generated in R version 4.3.1 using the ggplot2 version 2.1.0 and sf version 1.0-14 packages.

@@

bec_tab %>%
  gt() %>%
  cols_label(
    group = "BEC Group",
    zone = "BEC Zone",
    zone_nm = "Full Name",
    ppt = "Average Monthly Precipitation (mm/yr)",
    tmax = "Average Monthly Max Temperature (°C)",
    tmin = "Average Monthly Min Temperature (°C)",
    n = "n"
  ) %>%
  tab_header(title = "Biogeoclimatic Ecosystem Classification (BEC) Zones, their aggregated biomes, and their monthly average climate values for precipitation, maximum temperature, and minimum temperature. Climate data from Wang et al. 2016 averaged for 1991-2020.")
```

@@

The forest structural attributes (canopy height, canopy cover,
structural complexity \[coefficient of variation of height returns\],
basal area, aboveground biomass, and gross stem volume) were derived
from the 30-m spatial resolution dataset generated by Matasci et al.
[@matasci2018; @matasci2018b]. In brief, the method extends structural
attributes derived from a set of lidar acquisitions and field plots
across Canada. These data were generated for the treed pixels of Canada
using the BAP Landsat surface reflectance composite and auxiliary data
(i.e., topography, geography) using a k-Nearest Neighbour imputation
[@matasci2018; @matasci2018b]. Accuracy metrics for imputed forest
structural attributes range from an RMSE of 29.7% (structural
complexity) to 82.3% (gross stem volume) and *R^2^* ranging from 0.125
(structural complexity) to 0.712 (gross stem volume) [@matasci2018;
@matasci2018b].

@@

To obtain representative sample units of the EBV metrics across the
study area, we implemented stratified random sampling across the union
of BEC zones (Table 1) and forest types (i.e., coniferous, broadleaf,
mixed wood, and wetland treed; see @sec-study-area) of BC. Sample units
were restricted to forested pixels that were surrounded by the same
forest type to reduce the chance of sampling mixed pixels. We also
required that sample units were selected from a homogeneous area, to
reduce uncertainty associated with the input datasets [@shang2020];
homogeneous areas were defined as 3x3-pixel windows with a coefficient
of variation for canopy cover and canopy height below 0.5. A minimum
sampling distance of 1-km was implemented to reduce the effects of
spatial autocorrelation. Sample units that had been disturbed in the
last 35 years were discarded by using the disturbance mask generated for
the forested ecosystems of Canada [@hermosilla2016]. A maximum of 500
sample units were selected from each stratum; in the strata where this
was not achieved, all sampling units meeting the above restrictions were
selected. Sampling was conducted in R version 4.2.2 [@R-base] using the
**sgsR** package [@R-sgsR]. Neighbourhood analyses for the land cover
classes and coefficient of variations of canopy height and cover were
calculated in Python version 3.9.

@@

```{r}
#| label: fig-fcc
#| fig-cap: False colour maps by Biogeoclimatic Ecosystem Classification (BEC) zone of axis loadings for the first redundancy analysis (RDA) axis (A and B) and second RDA axis (C and D). Colour values are normalized to the maximum loading of each variable. A and C show axis loadings for canopy height (CH; red), canopy cover (CC; green) and structural complexity (SC; blue). B and D show axis loadings for basal area (BA; red), total aboveground biomass (AB; green) and gross stem volume (GSV; blue). Black indicates no significant loadings, while greyscale colours indicate similar loadings across the three variables. Annotated colour wheels for each set of variables also shown. Maps were generated in R version 4.3.1 using the ggplot2 version 2.1.0 and sf version 1.0-14 packages.

@@

BC shows varying forest structural drivers of the DHIs depending on the
geographic location (@fig-fcc). The interior of BC is drier and warmer,
and also has productivity being strongly driven by canopy cover, while
the Coastal BEC zone group has very low canopy cover loadings and high
structural complexity loadings (@fig-radar). The climate of these
regions could be influencing the DHIs [@perez2016], leading to varying
loadings strengths depending on the distance to the coast. For example,
water availability has been shown to be a driver of canopy height
[@klein2015]. In these wetter regions such as the Coastal group,
structural variability is more likely to be present in the vertical,
rather than horizontal direction, especially given the relationship
between canopy height and structural complexity [@atkins2022]. Further,
the weakest relationships between structure and function were found in
the Coastal group of BEC zones (@fig-fracts-bar), which may be due to
the consistently high amounts of canopy cover, and lower amounts of snow
found in these zones leading to potentially saturated NDVI values
throughout the year. These consistent forest structural metrics --- such
as the high canopy cover values found in the Coastal BEC zones --- may
also lead to lower amounts of variation explained by these metrics,
leading to them having smaller loadings, and other metrics with higher
variation in their values, such as structural complexity, having larger
loadings (@fig-radar). These complex linkages between climate, forest
structure, and forest productivity become further complicated by the
extreme environmental and topographic gradients within BC's ecosystems
[@pojar1987; @valentine1978soil], which may warrant study into the scale
of these interactions, with a finer (or non-climate based)
ecoregionalization potentially revealing different relationships between
ecosystem structure and function.
